Graph Topological Aspects of Granger Causal Network Learning\\ \large Supplementary Material

### 2019-07-18 paper_theory/proofs.tex
corrected proof of Prop 4.

  $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(b) \Rightarrow (c)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$.  The result follows since
  $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result of the equivalence
  in Definition \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(d) \implies (a)$ follows directly from the Definition.
\end{proof}

@@

@@

\begin{proof}
  We will prove by way of contradiction.  To this end, suppose that
  $j$ is a node such that: $(a)$ $j \not \in \anc{i}$ and $(b)$ for
  every $k \in \anc{i} \cap \anc{j}$ every
  $k \rightarrow \cdots \rightarrow j$ path contains $i$.

@@

  We notice that for any $u$ having the properties $(a), (b)$ above,
  we must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ (and
  $w \in \anc{i} \cap \anc{u}$, since $k \in \pa{u}$) such that
  $i \not \in \anc{k}$ and therefore there must be a path
  $\gcgpath{w}{k} \rightarrow u$ which does not contain $i$, contradicting
  property $(b)$.

  Using this partition, and in a similar fashion to the proof of
  Proposition \ref{prop:ancestor_uncorrelated}, we will expand
  $x_j(t)$ in terms of it's parents, and recursively expand nodes in
  $C_1$ until reaching $i$.  For the first step equation
  \eqref{eqn:parent_expansion} gives us:

  \begin{equation}
    \label{eqn:xj_partition_expansion}
    x_j(t) = \widetilde{\A}_{jj}(z) \Big(v_j(t) + \sum_{k \in C_0(j)} \B_{ik}(z)x_k(t) + \sum_{k \in C_1(j)}\B_{ik}(z)x_k(t)\Big).
  \end{equation}

  Using this representation we choose an arbitrary element of
  $\H_{t - 1}^j$ and represent it as the action of a strictly causal
  filter $\Phi(z)$, i.e.  $\Phi(z) x_j(t) \in \H_{t - 1}^{(j)}$ and show that

  \begin{equation}
    \label{eqn:sufficient_inner_prod}
    \inner{x_i(t)}{\widetilde{\A}_{jj}(z)\Phi(z)x_j(t) - \linE{\widetilde{\A}_{jj}(z)\Phi(z)x_j(t)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation}

  which will imply (by Theorem
  \ref{thm:granger_causality_equivalences}) that $j \npwgc i$.

  Substituting (\ref{eqn:xj_partition_expansion}) into
  (\ref{eqn:sufficient_inner_prod}) we have cancellation of all terms
  in which $k \in C_0(j)$ by Proposition \ref{prop:ancestor_properties},
  and cancellation of the term involving $v_j(t)$ by Lemma \ref{lem:vj_perp}.

  We are therefore left with

  \begin{equation}
    \label{eqn:refwornv}  % An arbitrary code for reference in just one place
    \inner{x_i(t)}{\sum_{k \in C_1(j)} \big(\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t) - \linE{\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t)}{\H_{t - 1}^i}\big)}.
  \end{equation}

  Since each such $k$ inherits the needed properties $(a), (b)$ we can
  again (as in Proposition \ref{prop:ancestor_properties}) recursively
  expand this expression until we reach $i$, obtaining a strictly
  causal $\Psi(z)$ such that Expression \eqref{eqn:refwornv} is equal to

  \begin{equation}
    \inner{x_i(t)}{\Psi(z)x_i(t) - \linE{\Psi(z)x_i(t)}{\H_{t - 1}^i}} = 0.
  \end{equation}
\end{proof}

>>>>

  $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(b) \Rightarrow (c)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{-j}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$.  The result follows since
  $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result of the equivalence
  in Definition \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(d) \implies (a)$ follows directly from the Definition.
\end{proof}

@@

\begin{remark}
  In order to prove Proposition \ref{prop:ancestor_properties} we
  require some additional notation, as well as another representation
  theorem.  The difficulty addressed by the following Definition
  \ref{def:scc} and Lemma \ref{lem:scc_expansion} is that in the
  representation of $x_j(t)$ in terms of it's parents (i.e. Equation
  \eqref{eqn:parent_expansion})

  \[
    x_i(t) = v_i(t) + \B_{ii}(z)x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t),
  \]

  the filter $\B_{ii}(z)$ need not be stable.  That is, the inverse
  filter $(1 - \B_{ii}(z))^{-1}$ need not exist.  An example of this
  issue is furnished by

  \[
    \B(z) = \left[
    \begin{array}{cc}
      \rho & -a\\
      a & 0\\
    \end{array}
  \right] z^{-1},
  \]

  for which, depending on the value of $a$, may still be stable even
  if $|\rho| > 1$.  This implies that it is not always possible to
  represent $x_i(t)$ in terms of $v_i(t)$ and $x_k(t), k \in \pa{i}$
  alone, i.e. as

  \[
    x_i(t) = (1 - \B_{ii}(z))^{-1}\big(v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\big).
  \]

  The difficulty presented by the non-existence of such a
  representation may become apparent upon studying the proof of
  Proposition \ref{prop:ancestor_properties}.
\end{remark}

\begin{definition}[Strongly Connected Components]
  \label{def:scc}
  In a graph $\gcg$, the subset $S \subseteq [n]$ is \textit{strongly
    connected} if $\forall i, j \in S$, $i \in \anc{j}$ and
  $j \in \anc{i}$.  We will denote by $S(j)$ (which may be the
  singleton $\{j\}$) the largest strongly connected component (SCC)
  containing $j$.  We will denote $x_{S(j)}(t)$ to be the vector of
  processes

  \[
    x_{S(j)}(t) = \big(x_s(t)\ |\ s \in S(j)\big),
  \]

  whose indices are sorted in ascending order.  Similarly, the
  sub-filter of $\B(z)$ acting on $x_{S(j)}(t)$ will be denoted
  $\B_{S(j)}(z)$.
\end{definition}

\begin{lemma}[Expansion in SCCs]
  \label{lem:scc_expansion}
  Given some $j \in [n]$, the process $x_{S(j)}(t)$ can be represented
  by
% {\substack{k \in \pa{a_0} \\ k \ne a_1}}
  \begin{equation}
    \label{eqn:scc_parent_expansion}
    x_{S(j)}(t) = v_{S(j)}(t) + \B_{S(j)}(z)x_{S(j)}(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} B_{sk}(z)x_k(t)e_s^{S(j)},
  \end{equation}

  where $e_s^{S(j)}$ denotes the length $|S(j)|$ canonical basis
  vector with a $1$ in the component corresponding to $x_s$ in the
  vector $x_{S(j)}$, and the summation is a double sum on $s$ and $k$.

  Moreover, the filter $\B(z)$ is stable with $I - \B_{S(j)}(z)$
  invertible:

  \begin{equation}
    \label{eqn:scc_inversion}
    (I - \B_{S(j)}(z))^{-1} = \sum_{k = 0}^\infty \B_{S(j)}(z)^k,
  \end{equation}

  therefore

  \begin{equation}
    \label{eqn:scc_parent_expansion_inverted}
    x_{S(j)}(t) = (I - \B_{S(j)}(z))^{-1} \big(v_{S(j)}(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} B_{sk}(z)x_k(t)e_s^{S(j)}\big),
  \end{equation}

\end{lemma}
\begin{proof}
  The representation of Equation \eqref{eqn:scc_parent_expansion}
  follows directly from the $\VAR$ representation of $x(t)$ (i.e. Equation
  \eqref{eqn:ar_representation})

  \[
    x(t) = \B(z)x(t) + v(t).
  \]

  \hl{TODO: show stability}
  % For the stability of $\B_{S(j)}(z)$ we must show that
  % $\forall z: |z^{-1}| \le 1$ we have $$||\B_{S(j)}(z)||_2 < 1.$$
\end{proof}

@@

\begin{proof}
  We will prove by way of contradiction.  To this end, suppose that
  $j$ is a node such that:

  \begin{enumerate}[label=(\alph*)]
    \item{$j \not \in \anc{i}$}
    \item{every $k \in \anc{i} \cap \anc{j}$ every
        $k \rightarrow \cdots \rightarrow j$ path contains $i$.}
  \end{enumerate}

@@

  We notice that for any $u$ having the properties $(a), (b)$ above,
  we must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ (and
  $w \in \anc{i} \cap \anc{u}$, since $k \in \pa{u}$) such that
  $i \not \in \anc{k}$ and therefore there must be a path
  $\gcgpath{w}{k} \rightarrow u$ which does not contain $i$,
  contradicting property $(b)$.  Moreover, for any
  $u \in \big(\anc{j} \setminus \{i\}\big)$ and $k \in C_0(u)$,
  Proposition \ref{prop:ancestor_uncorrelated} shows that
  $H_t^i \perp \H_{t - 1}^j | \H_{t - 1}^i$.

  In order to establish $j \npwgc i$, choose an arbitrary element of
  $\H_{t - 1}^j$ and represent it via the action of a strictly causal
  filter $\Phi(z)$, i.e.  $\Phi(z) x_j(t) \in \H_{t - 1}^{(j)}$, by
  Theorem \ref{thm:granger_causality_equivalences} it suffices to show
  that

  \begin{equation}
    \label{eqn:sufficient_inner_prod}
    \inner{x_i(t)}{\Phi(z)x_j(t) - \linE{\Phi(z)x_j(t)}{\H_{t - 1}^{(i)}}} = 0.
  \end{equation}

  Denote $e_j \defeq e_j^{S(j)}$, we can write
  $x_j(t) = e_j^\T x_{S(j)}(t)$, and therefore from Equation
  \eqref{eqn:scc_parent_expansion_inverted} there exist strictly
  causal filters $\Gamma_s(z)$ and $\Lambda_{sk}(z)$ (defined for ease
  of notation) such that

  \[
    x_j(t) = \sum_{s \in S(j)} \Gamma_s(z)v_s(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} \Lambda_{sk}(z)x_k(t).
  \]

  When we substitute this expression into the left hand side of
  Equation \eqref{eqn:sufficient_inner_prod}, we may cancel each term
  involving $v_s$ by Lemma \ref{lem:vj_perp}, and each $k \in C_0(s)$
  by our earlier argument, leaving us with

  \[
    \sum_{\substack{s \in S(j) \\ k \in C_1(s) \cap S(j)^\c}}\inner{x_i(t)}{\Phi(z)\Lambda_{sk}(z)x_k(t) - \linE{\Phi(z)\Lambda_{sk}(z)x_k(t)}{\H_{t - 1}^{(i)}}}.
  \]

  Since each $k \in C_1(s)$ with $k \ne i$ inherits properties $(a)$
  and $(b)$ above, we can recursively expand each $x_k$ of the above
  summation until reaching $k = i$ (which is garaunteed to terminate
  due to the definition of $C_1(u)$) which leaves us with some
  strictly causal filter $F(z)$ such that the left hand side of
  Equation \eqref{eqn:sufficient_inner_prod} is equal to

  \[
    \inner{x_i(t)}{\Phi(z)F(z)x_i(t) - \linE{\Phi(z)F(z)x_i(t)}{\H_{t - 1}^{(i)}}},
  \]

  and this is $0$ since $\Phi(z)F(z)x_i(t) \in \H_{t - 1}^i$.

  % Using the above partition of $\pa{j}$, we can expand $x_j(t)$ in
  % terms of it's parents, and recursively expand nodes in $C_1$
  % (cancelling all nodes in $C_0$, and recalling that
  % $C_2 = \emptyset$) until we reach $i$.

  % Using this partition, we will expand $x_j(t)$ in terms of it's
  % parents, and recursively expand nodes in $C_1$ until reaching $i$.
  % For the first step equation \eqref{eqn:parent_expansion} gives us:

  % \begin{equation}
  %   \label{eqn:xj_partition_expansion}
  %   x_j(t) = \widetilde{\A}_{jj}(z) \Big(v_j(t) + \sum_{k \in C_0(j)} \B_{ik}(z)x_k(t) + \sum_{k \in C_1(j)}\B_{ik}(z)x_k(t)\Big).
  % \end{equation}

  % Substituting (\ref{eqn:xj_partition_expansion}) into
  % (\ref{eqn:sufficient_inner_prod}) we have cancellation of all terms
  % in which $k \in C_0(j)$ by Proposition \ref{prop:ancestor_properties},
  % and cancellation of the term involving $v_j(t)$ by Lemma \ref{lem:vj_perp}.

  % We are therefore left with

  % \begin{equation}
  %   \label{eqn:refwornv}  % An arbitrary code for reference in just one place
  %   \inner{x_i(t)}{\sum_{k \in C_1(j)} \big(\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t) - \linE{\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t)}{\H_{t - 1}^i}\big)}.
  % \end{equation}

  % Since each such $k$ inherits the needed properties $(a), (b)$ we can
  % again (as in Proposition \ref{prop:ancestor_properties}) recursively
  % expand this expression until we reach $i$, obtaining a strictly
  % causal $\Psi(z)$ such that Expression \eqref{eqn:refwornv} is equal to

  % \begin{equation}
  %   \inner{x_i(t)}{\Psi(z)x_i(t) - \linE{\Psi(z)x_i(t)}{\H_{t - 1}^i}} = 0.
  % \end{equation}
\end{proof}
